---
title: "New York City Stop-Question-Frisk Analysis"
author: "Jacky Lee, Tyler Sam, Sam Tan, Julianne Lin"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(eval=TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
library(kernlab)
library(reshape2)
library(ggplot2)
data(spam)
sqf2010 = read.csv("2010_sqf_m35.csv")
sqf2015 = read.csv("2015_sqf_m35.csv")
```

**The report should have the following structure:**

1. Introduction (overview in your own words of the data; summary of your team's chosen question, why it's important, and what you found; roadmap for the remainder of the report.)

2. Responses to required questions: Please list each question/answer separately, but write formal responses to the questions.

3. Analysis of your chosen questions: Be sure to describe your questions and why they are interesting/important, and provide suitable graphical and statistical analysis of the data to answer the questions.

4. Conclusions and Recommendations

5. Bibliography if you use any external sources

6. Appendices as necessary

**Your report will be evaluated based on the following criteria:**
- Writing: Structure of report; clarity of writing; quality of editing and polishing the document

- Correctness of methodology: Appropriate graphical and statistical techniques are used to answer the questions

- Validity of interpretation: The results are interpreted correctly

- Significance: The team-chosen question(s) leverage the available data to provide insight into the stop-question-frisk policy.

```{r}
#1.
testedHispanic = length(sqf2010$race[sqf2010$race=='WHITE-HISPANIC' | sqf2010$race=='BLACK-HISPANIC'])
testedAsian = length(sqf2010$race[sqf2010$race=='ASIAN/PACIFIC ISLANDER'])
trueHispanic = testedHispanic*19/18 - testedAsian/18
trueAsian = testedAsian*19/18 - testedHispanic/18
total = testedHispanic + testedAsian

probHispanic = trueHispanic / total
probAsian = trueAsian / total
probActualHispanic = 0.95*trueHispanic / testedHispanic
probActualAsian = 0.95*trueAsian / testedAsian
```

```{r}
#2.
par(mfrow=c(2,3))
races = c('AMERICAN INDIAN/ALASKAN NATIVE','ASIAN/PACIFIC ISLANDER','BLACK','BLACK-HISPANIC','WHITE','WHITE-HISPANIC')
city = c('BRONX','BROOKLYN','MANHATTAN','QUEENS','STATEN IS')

analyze = function(dataSet,title) {
    frisked = c()
    data = c()
    for (i in races) {
        data = append(data,dataSet$frisked[dataSet$race == i])
        frisked = append(frisked,sum(data) / length(data))
    }
    barplot(frisked,main=title,xlab='Race',ylab='Proportion Frisked',names.arg=races,cex.names=0.4,las=2)
    
    #t.test(data[races == "WHITE"],data[races == "BLACK"])
    #t.test(data[races == races[2]],data[races == races[3]])
}

analyze(sqf2010,'Proportion of People Frisked By Race')
for (i in city) {
    analyze(sqf2010[sqf2010$city==i,],i)
}
queens = sqf2010[sqf2010$city=='QUEENS',]
bronx = sqf2010[sqf2010$city=='BRONX',]


```

```{r}
#3.
agesM = subset(sqf2010, sqf2010$age>5&sqf2010$age<98&sqf2010$sex =="M")
agesF = subset(sqf2010, sqf2010$age>5&sqf2010$age<98&sqf2010$sex =="F")
qqnorm(agesM$age)
qqnorm(agesF$age)
# both subsets of the age appear normal
qqplot(agesM$age, agesF$age)
# since the qqplot is a line, the distributions are the same.

agesStop = sqf2010$age[sqf2010$age<98&sqf2010$age>5]
qqplot(agesM$age, agesStop)
# since the qqplot is a line, the distributions are the same.
```

```{r}
#4.
sample_size = 22563
# the sample size of the 2015 data set is 22563
sample_mean <- mean(sqf2015$frisked)
z_critical <- qnorm(0.975)
pop_stdev <- sd(sqf2015$frisked)
margin_of_error <- z_critical * (pop_stdev / sqrt(sample_size))
confidence_interval <- c(sample_mean - margin_of_error,
                         sample_mean + margin_of_error)
print("Confidence interval:")
confidence_interval

# Confidence Interval for 2015 rates for being frisked(0.6700898 0.6823013)

summary(sqf2015)
sample_size = 639
# the sample size of the 2015 data set is 22563
sample_noId = subset(sqf2015, sqf2015$typeofid =="REFUSED" )
sample_mean <- mean(sample_noId$frisked)
z_critical <- qnorm(0.975)
pop_stdev <- sd(sample_noId$frisked)
margin_of_error <- z_critical * (pop_stdev / sqrt(sample_size))
confidence_interval <- c(sample_mean - margin_of_error,
                         sample_mean + margin_of_error)
print("Confidence interval:")
confidence_interval
# Confidence Interval for 2015 rates for being frisked given that
# one refused to show ID(0.5724871 0.6481702)
```

```{r}
#5.
m_best = step(lm(formula = perstop ~ arstmade + searched + inside + sumissue + frisked + weap + contrabn + radio + pf, data = sqf2010[sqf2010$perstop<30,]))
plot(m_best$fitted.values, m_best$residuals)
qqplot(m_best$fitted.values, m_best$residuals)

# m_full = lm(complied ~ pop + hdi + dem + internet + freepress, data = goog_sub)
# m_best = step(m_full)
# plot(m_best$fitted.values, m_best$residuals)
```

Self-generated questions:

According to an analysis of the Stop and Frisk measure between 2002 and 2013 done by the New York Civil Liberties Union, young black and Latino men were stopped disproportionately by a huge margin. It claimed that black and Latino males between the ages of 14 and 24 accounted for 41 percent of stops between 2003 and 2013, even though they were only 4.7 of the total city population. Furthermore, according to their analysis, nearly 90 percent of young black and Latino men stopped were innocent.
Verify if these trends continued in the year 2015.
Link: http://www.nyclu.org/files/publications/stopandfrisk_briefer_2002-2013_final.pdf

According to this analysis of race and crime in Chicago, Black people are 85\% more likely than Whites to commit a violent crime compared to whites, while Latino people are 10\% less likely than Whites. Compare this data to the proportion of Black and Latino people who were stopped. Does the data support stopping Blacks and Latinos at different rates?
Link: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1449156/