---
title: "New York City Stop-Question-Frisk Analysis"
author: "Jacky Lee, Tyler Sam, Sam Tan, Julianne Lin"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(eval=TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
library(kernlab)
library(reshape2)
library(ggplot2)
data(spam)
sqf2010 = read.csv("2010_sqf_m35.csv")
sqf2015 = read.csv("2015_sqf_m35.csv")
```

**The report should have the following structure:**

1. Introduction (overview in your own words of the data; summary of your team's chosen question, why it's important, and what you found; roadmap for the remainder of the report.)

2. Responses to required questions: Please list each question/answer separately, but write formal responses to the questions.

3. Analysis of your chosen questions: Be sure to describe your questions and why they are interesting/important, and provide suitable graphical and statistical analysis of the data to answer the questions.

4. Conclusions and Recommendations

5. Bibliography if you use any external sources

6. Appendices as necessary

**Your report will be evaluated based on the following criteria:**
- Writing: Structure of report; clarity of writing; quality of editing and polishing the document

- Correctness of methodology: Appropriate graphical and statistical techniques are used to answer the questions

- Validity of interpretation: The results are interpreted correctly

- Significance: The team-chosen question(s) leverage the available data to provide insight into the stop-question-frisk policy.

(No Intro yet)

1. The question asks for the conditional probability that you are actually a certain race given you are identified as that race, P(actual | tested) where tested implies identification. We are given the probability of correct identification. Thus, given the data from SQF2010, we are able to set up a system of equations to solve for the number of true Asians and true Hispanics. We used this information to calculate the probability of actually being Asian/Hispanic given the identification. This calculation shows us that the actual proportions of Asians and Hispanics is different from what the data claims it to be.

```{r}
testedHispanic = length(sqf2010$race[sqf2010$race=='WHITE-HISPANIC' | sqf2010$race=='BLACK-HISPANIC'])
testedAsian = length(sqf2010$race[sqf2010$race=='ASIAN/PACIFIC ISLANDER'])
trueHispanic = testedHispanic*19/18 - testedAsian/18
trueAsian = testedAsian*19/18 - testedHispanic/18
total = testedHispanic + testedAsian

probHispanic = trueHispanic / total
probAsian = trueAsian / total
probActualHispanic = 0.95*trueHispanic / testedHispanic
probActualAsian = 0.95*trueAsian / testedAsian
```

2. 

```{r}
par(mfrow=c(2,3))
races = c('AMERICAN INDIAN/ALASKAN NATIVE','ASIAN/PACIFIC ISLANDER','BLACK','BLACK-HISPANIC','WHITE','WHITE-HISPANIC')
city = c('BRONX','BROOKLYN','MANHATTAN','QUEENS','STATEN IS')

analyze = function(dataSet,title) {
    frisked = c()
    data = c()
    for (i in races) {
        data = append(data,dataSet$frisked[dataSet$race==i])
        frisked = append(frisked,sum(data)/length(data))
    }
    barplot(frisked,main=title,xlab='Race',ylab='Proportion Frisked',names.arg=races,cex.names=0.4,las=2)
}

analyze(sqf2010,'Proportion of People Frisked By Race')
for (i in city) {
    analyze(sqf2010[sqf2010$city==i,],i)
}
queens = sqf2010[sqf2010$city=='QUEENS',]
bronx = sqf2010[sqf2010$city=='BRONX',]

sqf2010$raceWhite = relevel(sqf2010$race,'WHITE')
summary(lm(frisked ~ raceWhite,data=sqf2010))

# most significant differences
summary(lm(formula=frisked ~ raceWhite,data=subset(sqf2010,city=='STATEN IS')))

# most similar rates
summary(lm(formula=frisked ~ raceWhite,data=subset(sqf2010,city=='QUEENS')))
```

```{r,eval=FALSE}
summary(lm(formula=frisked ~ raceWhite,data=subset(sqf2010,city=='BRONX')))
summary(lm(formula=frisked ~ raceWhite,data=subset(sqf2010,city=='BROOKLYN')))
summary(lm(formula=frisked ~ raceWhite,data=subset(sqf2010,city=='MANHATTAN')))
```

3. We analyzed the distribution of the ages of male suspects and female suspects individually and then compare them with each other. First, we took out extraneous data by creating subsets that exclude ages below 5 and above 98. To determine their individual distributions, we used qqnorm to check for normality and concluded that both male and female suspects' ages are distributed normally. To analyze both distributions in relation with each other, we used a qqplot with the subsets of the ages of male and female suspects. The qqplot appears to be linear, thus we concluded that the distributions are the same. Later, we analyzed the age distribution for male suspects in relation to the age distribution of the entire population of suspects stopped in 2010. Again, we created a subset of the population of stopped suspects by limitingthe age to between 5 and 98. We then compared using the qqplot command using the arguments of the male subset from earlier and the total populatio subset we created. Again, qqplot shows a linear relationship so we conclude the distributions are the same. In conclusion, both distributions of the ages of male and female suspects in 2010 are normally distributed. Both these age distributions are the same, and the the male age distribution and the total stopped suspect distribution are also the same.

```{r}
agesM = subset(sqf2010,sqf2010$age>5 & sqf2010$age<98 & sqf2010$sex=="M")
agesF = subset(sqf2010,sqf2010$age>5 & sqf2010$age<98 & sqf2010$sex=="F")
agesStop = sqf2010$age[sqf2010$age>5 & sqf2010$age<98]
par(mfrow=c(2,2))
qqnorm(agesM$age)
qqline(agesM$age)
qqnorm(agesF$age)
qqline(agesF$age)
qqplot(agesM$age, agesF$age)
qqplot(agesM$age, agesStop)
```

4. To analyze any differences between the probability of being frisked in general and the probability of being frisked when not providing identification, we first calculated the confidence intervals for both. To calculate the CI for the general sample of frisked suspects, we first found the sample size, mean, critical z-value, and standard deviation. Then using that information, we found the margin of error from the sample mean and used that to calculate the confidence interval. The interval for 2015 rate of being frisked is (0.6700898, 0.6823013). Thus we have 95% confidence that the probability of being frisked is between 0.67 and 0.68, roughly speaking. Similarly, we calculated the confidence interval of the sample of frisked suspects who didn't provide identification. In this case, our sample is the subset of 2015 suspects whose identification is listed as 'REFUSED'. Using the same process we found the CI of 95% confidence for the sample of 2015 frisked suspects who didn't provide ID. Our confidence interval came out as (0.5724871, 0.6481702). This means that the probability of being frisked having refused identification is between 0.57 and 0.65. In conclusion, these two samples have a significant difference in probability of being frisked. The confidence intervals don't overlap, which means that the the probability of anyone from the general population of being frisked is outside the 95% confidence interval of that of those who refused identification. Basically, we conclude from our data that we have significant enough data to say that in 2015 there is a smaller probability of being frisked when identification is refused.

```{r}
sample_size = length(sqf2015$frisked)
sample_mean = mean(sqf2015$frisked)
z_critical = qnorm(0.975)
pop_stdev = sd(sqf2015$frisked)
margin_of_error = z_critical * (pop_stdev / sqrt(sample_size))
confidence_interval = c(sample_mean - margin_of_error,sample_mean + margin_of_error)
print("Confidence interval:")
confidence_interval

# Confidence Interval for 2015 rates for being frisked(0.6700898 0.6823013)

summary(sqf2015)
sample_size = 639
# the sample size of the 2015 data set is 22563
sample_noId = subset(sqf2015, sqf2015$typeofid =="REFUSED" )
sample_mean <- mean(sample_noId$frisked)
z_critical <- qnorm(0.975)
pop_stdev <- sd(sample_noId$frisked)
margin_of_error <- z_critical * (pop_stdev / sqrt(sample_size))
confidence_interval <- c(sample_mean - margin_of_error,
                         sample_mean + margin_of_error)
print("Confidence interval:")
confidence_interval
# Confidence Interval for 2015 rates for being frisked given that
# one refused to show ID(0.5724871 0.6481702)
```

5. 

```{r}
#5.
m_best = step(lm(formula = perstop ~ arstmade + searched + inside + sumissue + frisked + weap + contrabn + radio + pf, data = sqf2010[sqf2010$perstop<30,]))
plot(m_best$fitted.values, m_best$residuals)
qqplot(m_best$fitted.values, m_best$residuals)

# m_full = lm(complied ~ pop + hdi + dem + internet + freepress, data = goog_sub)
# m_best = step(m_full)
# plot(m_best$fitted.values, m_best$residuals)
```

Self-generated questions:

According to an analysis of the Stop and Frisk measure between 2002 and 2013 done by the New York Civil Liberties Union, young black and Latino men were stopped disproportionately by a huge margin. It claimed that:

(1) Black and Latino males between the ages of 14 and 24 accounted for 41 percent of stops between 2003 and 2013, even though they were only 4.7 of the total city population. (2) Furthermore, according to their analysis, 90 percent of young black and Latino men stopped should not have been stopped because they were not arrested.
Verify that these trends continued in the year 2015.
Link: http://www.nyclu.org/files/publications/stopandfrisk_briefer_2002-2013_final.pdf

1. For the first part of this question, we are attempting to see if the trend from the 2002-2013 analysis that Black and Latino males between 14 and 24 accounted for 41% of stops is true for 2015. This analysis is significant because the analysis for 2002-2013 shows that young Black and Latino males make up almost half of the stops despite the small percentage of the population they represent. This information shows potential racial discrimination in "suspects". If this same trend is identified in the 2015 data, that shows that between 2013 and 2015, racial discrimination has not lessened. If the trend is not seen in 2015, we can identify racial discrimination towards Black and Latino male suspects has either lessened or increased. 

To identify the presence of this trend in the 2015 suspect data, we created a subset for the two samples. First we created a subset of the respective races (Black-Hispanic and White-Hispanic for the Hispanic sample, and Black for the Black sample) and the ages trimmed to greater than 13 and less than 25 to comply with the sample of Blacks and Hispanics of age 14-24. Then, we further narrowed the subset by creating another subset of the previous by specifying sex as male. We found the population of this sample by finding the dimensions of this subset and got 8749. Using dimension again, we also found the population of the entire sample of all suspects, which was 22563. In this case, our hypothesis is that the mean of the 2015 data for stopped young Black and Hispanic males is equal to 41%. The alternative hypothesis is that the 2015 mean is NOT 41%. To disprove or fail to disprove this hypothesis, we found the 95% confidence interval using both the populations we calculated above (8749 and 22564 respectively). The confidence interval we found is (0.381, 0.394). 41% or 0.41 is outside the 95% interval, so we can disprove our initial hypothesis. We can say with 95% confidence that the percentage of stopped suspects that are Black and Hispanic males between 14 and 24 is between 38% and 39%. Thus we can conclude that between 2013 and 2015, the proportion of males who are Black and Hispanic between 14 and 24 decreased. This is important because we can then speculate as to whether discrimination against Black and Hispanic suspects has lessened. 

2. In the second part of this question, we trying to identify how many of the stopped Black and Hispanic males were actually not arrested in 2015. In the 2003-2013 analysis shows that 90% of young Black and Hispanic males stopped suspects were not arrested. Again, this furthers the speculation that there is racial discrimination against Black and Hispanic suspects because out of those stopped, a high percentage were actually not found to have enough evidence of crime for arrest. Our null hypothesis is that the percentage of non-arrested in 2015 is also 90%, and our alternate hypothesis is that the percentage is NOT 90%. Again, this is significant because proving or disproving the null hypothesis will allow us to concluded whether or not there is significant change from 2002-2013 and thus see if potential racial discrimination has increased or decreased or remained the same.

To test our hypothesis, we first created a subset of the male Black/Hispanic between 14 and 24 sample we created in part 1. We narrowed down that sample by specifying "arrest made" as negative, or 0. The number of non-arrested Black/Hispanic male suspects between 14 and 24 is 7761. The 95% confidence interval using that number and the number of total Black/Hispanic young, male suspects stopped (8749 found in part 1) is (0.880, 0.894). In this situation, we reject the null hypothesis because 90% (or 0.9) is not within the confidence interval we found for the 2015 data. We can say with 95% confidence that the percentage of non-arrested from the set of Black/Hispanic stopped male suspects (between 14-24) decreased from 2002-2013. This suggests that more of the stopped suspects were found to be guilty enough for arrest. We can speculate that racial discrimination decreased because more of the stopped suspects should have been stopped and so less stopped suspects were stopped without "reason". 

***(idk if I can say this question is significant because of racial discrimination tho because u could also argue that the percentage decreased because police are arresting more innocent people so racial discrimination actual increased)

```{r}
# Question 1
ages1424BL= subset(sqf2015, sqf2015$age>13 & sqf2015$age<25 & (sqf2015$race =="BLACK" | sqf2015$race == "WHITE-HISPANIC" | sqf2015$race == "BLACK-HISPANIC"))
ages1424BLM = subset(ages1424BL, ages1424BL$sex == "M")
dim(ages1424BLM)
# the number of samples in sqf2015 given that one is male, black or hispanic, and between the ages 14 and 24 is 8749.
dim(sqf2015)
# the number of samples in sqf2015 is 22563
prop.test(8749, 22563, alternative = c("two.sided", "less", "greater"), conf.level = .95)
# the 95% confidence interval for this proportion is (.381, .394).
# Since .41 is not in our 95% confidence interval, we can conclude that the proportion of males who are black or latino between the ages 14 and 24 decreased in 2015 compared to the percentage in 2003 to 2013.


#Question 2
ages1424BLMI = subset(ages1424BLM, ages1424BL$arstmade == 0)
dim(ages1424BLMI)
# the number of samples in sqf2015 given that one is male, black or hispanic, between the ages 14 and 24, and not arrested is 7761
dim(ages1424BLM)
# the number of samples in sqf2015 given that one is male, black or hispanic, and between the ages 14 and 24 is 8749.
prop.test(7761, 8749, alternative = c("two.sided", "less", "greater"), conf.level = .95)
# the 95% confidence interval for this proportion is (.880, .894).
# Since .9 is not in our 95% confidence interval, we can conclude that the proportion of males who are black or latino, between the ages 14 and 24, and not-arrested decreased in 2015 compared to the percentage in 2003 to 2013.
```

According to this analysis of race and crime in Chicago, Black people are 85% more likely than Whites to commit a violent crime compared to whites, while Latino people are 10% less likely than Whites. Compare this data to the proportion of Black and Latino people who were stopped. (3)Does the data of sqf2015 support stopping blacks at higher rates compared to whites? (4)Does the data of sqf2015 support stopping latinos at lower rates compared?
Link: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1449156/

```{r}
# Question 3
blackArrest = subset(sqf2015, sqf2015$arstmade == 1 & sqf2015$race == "BLACK")
whiteArrest = subset(sqf2015,sqf2015$arstmade == 1 & sqf2015$race == "WHITE")

#proportion of blacks who were arrested out of all black individuals who were stopped by the police.
dim(blackArrest)
dim(subset(sqf2015, sqf2015$race == "BLACK"))

prop.test(1993, 11950, alternative = c("two.sided", "less", "greater"), conf.level = .95)
# the 95% confidence interval for this proportion is (.160, .174).

dim(whiteArrest)
dim(subset(sqf2015, sqf2015$race == "WHITE"))

prop.test(344, 2514, alternative = c("two.sided", "less", "greater"), conf.level = .95)
# the 95% confidence interval for this proportion is (.124, .151).
# Based on the results from the two proportion tests above, since the confidence intervals do not overlap, statistically speaking, it is justified for police to stop more blacks than whites. 

# Question 4
latinoArrest = subset(sqf2015, sqf2015$arstmade == 1 & (sqf2015$race == "BLACK-HISPANIC" | sqf2015$race == "WHITE-HISPANIC"))

#proportion of blacks who were arrested out of all black individuals who were stopped by the police.
dim(latinoArrest)
dim(subset(sqf2015, sqf2015$race == "BLACK-HISPANIC" | sqf2015$race == "WHITE-HISPANIC"))

prop.test(1423, 6499, alternative = c("two.sided", "less", "greater"), conf.level = .95)
# the 95% confidence interval for this proportion is (.209, .229).

dim(whiteArrest)
dim(subset(sqf2015, sqf2015$race == "WHITE"))

prop.test(344, 2514, alternative = c("two.sided", "less", "greater"), conf.level = .95)
# the 95% confidence interval for this proportion is (.124, .151).
# Based on the results from the two proportion tests above, since the confidence intervals do not overlap, statistically speaking, it is not justified for police to stop more whites than latinos. 
```

