---
title: "New York City Stop-Question-Frisk Analysis"
author: "Jacky Lee, Tyler Sam, Sam Tan, Julianne Lin"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(eval=TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
library(kernlab)
library(reshape2)
library(ggplot2)
data(spam)
sqf2010 = read.csv("2010_sqf_m35.csv")
sqf2015 = read.csv("2015_sqf_m35.csv")
```

**The report should have the following structure:**

1. Introduction (overview in your own words of the data; summary of your team's chosen question, why it's important, and what you found; roadmap for the remainder of the report.)

2. Responses to required questions: Please list each question/answer separately, but write formal responses to the questions.

3. Analysis of your chosen questions: Be sure to describe your questions and why they are interesting/important, and provide suitable graphical and statistical analysis of the data to answer the questions.

4. Conclusions and Recommendations

5. Bibliography if you use any external sources

6. Appendices as necessary

**Your report will be evaluated based on the following criteria:**
- Writing: Structure of report; clarity of writing; quality of editing and polishing the document

- Correctness of methodology: Appropriate graphical and statistical techniques are used to answer the questions

- Validity of interpretation: The results are interpreted correctly

- Significance: The team-chosen question(s) leverage the available data to provide insight into the stop-question-frisk policy.

```{r}
#1.
testedHispanic = length(sqf2010$race[sqf2010$race=='WHITE-HISPANIC' | sqf2010$race=='BLACK-HISPANIC'])
testedAsian = length(sqf2010$race[sqf2010$race=='ASIAN/PACIFIC ISLANDER'])
trueHispanic = testedHispanic*19/18 - testedAsian/18
trueAsian = testedAsian*19/18 - testedHispanic/18
total = testedHispanic + testedAsian

probHispanic = trueHispanic / total
probAsian = trueAsian / total
probActualHispanic = 0.95*trueHispanic / testedHispanic
probActualAsian = 0.95*trueAsian / testedAsian
```

```{r}
#2.
par(mfrow=c(2,3))
races = c('AMERICAN INDIAN/ALASKAN NATIVE','ASIAN/PACIFIC ISLANDER','BLACK','BLACK-HISPANIC','WHITE','WHITE-HISPANIC')
city = c('BRONX','BROOKLYN','MANHATTAN','QUEENS','STATEN IS')

analyze = function(dataSet,title) {
    frisked = c()
    data = c()
    for (i in races) {
        data = append(data,dataSet$frisked[dataSet$race == i])
        frisked = append(frisked,sum(data) / length(data))
    }
    barplot(frisked,main=title,xlab='Race',ylab='Proportion Frisked',names.arg=races,cex.names=0.4,las=2)
    
    #t.test(data[races == "WHITE"],data[races == "BLACK"])
    #t.test(data[races == races[2]],data[races == races[3]])
}

analyze(sqf2010,'Proportion of People Frisked By Race')
for (i in city) {
    analyze(sqf2010[sqf2010$city==i,],i)
}
queens = sqf2010[sqf2010$city=='QUEENS',]
bronx = sqf2010[sqf2010$city=='BRONX',]


```

```{r}
#3.
agesM = subset(sqf2010, sqf2010$age>5&sqf2010$age<98&sqf2010$sex =="M")
agesF = subset(sqf2010, sqf2010$age>5&sqf2010$age<98&sqf2010$sex =="F")
qqnorm(agesM$age)
qqnorm(agesF$age)
# both subsets of the age appear normal
qqplot(agesM$age, agesF$age)
# since the qqplot is a line, the distributions are the same.

agesStop = sqf2010$age[sqf2010$age<98&sqf2010$age>5]
qqplot(agesM$age, agesStop)
# since the qqplot is a line, the distributions are the same.
```

```{r}
#4.
sample_size = 22563
# the sample size of the 2015 data set is 22563
sample_mean <- mean(sqf2015$frisked)
z_critical <- qnorm(0.975)
pop_stdev <- sd(sqf2015$frisked)
margin_of_error <- z_critical * (pop_stdev / sqrt(sample_size))
confidence_interval <- c(sample_mean - margin_of_error,
                         sample_mean + margin_of_error)
print("Confidence interval:")
confidence_interval

# Confidence Interval for 2015 rates for being frisked(0.6700898 0.6823013)

summary(sqf2015)
sample_size = 639
# the sample size of the 2015 data set is 22563
sample_noId = subset(sqf2015, sqf2015$typeofid =="REFUSED" )
sample_mean <- mean(sample_noId$frisked)
z_critical <- qnorm(0.975)
pop_stdev <- sd(sample_noId$frisked)
margin_of_error <- z_critical * (pop_stdev / sqrt(sample_size))
confidence_interval <- c(sample_mean - margin_of_error,
                         sample_mean + margin_of_error)
print("Confidence interval:")
confidence_interval
# Confidence Interval for 2015 rates for being frisked given that
# one refused to show ID(0.5724871 0.6481702)
```

```{r}
#5.
m_best = step(lm(formula = perstop ~ arstmade + searched + inside + sumissue + frisked + weap + contrabn + radio + pf, data = sqf2010[sqf2010$perstop<30,]))
plot(m_best$fitted.values, m_best$residuals)
qqplot(m_best$fitted.values, m_best$residuals)

# m_full = lm(complied ~ pop + hdi + dem + internet + freepress, data = goog_sub)
# m_best = step(m_full)
# plot(m_best$fitted.values, m_best$residuals)
```
